---
title: 'International Trade: Assignment 1'
runningheader: 'Assignment 1'
author: |
  Carlos Lezama, Marco Hinojosa, \
  Luis Jiménez and Antonio Carbonell
date: 'October 27th, 2021'
output:
  tufte::tufte_handout:
    citation_package: natbib
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(tidyverse)
library(knitr)
library(expm)
library(matlib)
library(ggcorrplot)
library(mFilter)
library(zoo)

knitr::opts_chunk$set(
  cache.extra = packageVersion('tufte'),
  cache = FALSE,
  echo = FALSE,
  fig.fullwidth=TRUE,
  fig.height = 6,
  fig.pos = '!h',
  fig.width = 12,
  message = FALSE,
  out.extra = '',
  out.width = '100%',
  warning = FALSE
)

options(htmltools.dir.version = FALSE)
```

# Patterns of trade

## (a)

**i. and ii.**

```{r data.1}
# Data import
data.1 <- read.csv('data/world-bank.csv') %>%
  slice(-c(6:10))
data.1 <- data.1[-c(1:4)]

# Columns rename
names(data.1) <- c(1960:2020)

# Variables
gdp <- data.1[1, ]
merchExports <- data.1[2, ]
merchImports <- data.1[3, ]
goods_and_servicesExports <- data.1[4, ]
goods_and_servicesImports <- data.1[5, ]
```

```{r merchandise}
# Important merchandise ratios
merchExportsGDP <- t(merchExports / gdp)
merchImportsGDP <- t(merchImports / gdp)
merchGDP <- t((merchExports  + merchImports) / gdp)
merchandise <- data.frame(merchExportsGDP, merchImportsGDP, merchGDP)

# Mean before and after NAFTA
ratio1.before <- mean(merchGDP[as.character(c(1960:1993)), ])
ratio1.after <- mean(merchGDP[as.character(c(1994:2011)), ])

ratio2.before <- mean(merchExportsGDP[as.character(c(1960:1993)), ])
ratio2.after <- mean(merchExportsGDP[as.character(c(1994:2011)), ])

ratio3.before <- mean(merchImportsGDP[as.character(c(1960:1993)), ])
ratio3.after <- mean(merchImportsGDP[as.character(c(1994:2011)), ])

colours <-
  c(
    'Merchandise trade to GDP' = '#051c2c',
    'Merchandise exports to GDP' = '#1e40ca',
    'Merchandise imports to GDP' = '#00a2ed'
  )

# Plot
ggplot(merchandise) +
  geom_line(aes(
    x = as.numeric(row.names(merchandise)),
    y = merchGDP,
    colour = 'Merchandise trade to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(merchandise)),
    y = merchExportsGDP,
    colour = 'Merchandise exports to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(merchandise)),
    y = merchImportsGDP,
    colour = 'Merchandise imports to GDP'
  )) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio1.before,
      yend = ratio1.before,
      colour = 'Merchandise trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio1.after,
      yend = ratio1.after,
      colour = 'Merchandise trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio2.before,
      yend = ratio2.before,
      colour = 'Merchandise exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio2.after,
      yend = ratio2.after,
      colour = 'Merchandise exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio3.before,
      yend = ratio3.before,
      colour = 'Merchandise imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio3.after,
      yend = ratio3.after,
      colour = 'Merchandise imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(x = 'year', y = 'ratio') +
  scale_colour_manual(name = 'Ratios', values = colours) +
  theme_minimal()
```

* Merchandise trade to GDP ratio average before NAFTA: `r ratio1.before`
* Merchandise trade to GDP ratio average after NAFTA: `r ratio1.after`
* Merchandise exports to GDP ratio average before NAFTA: `r ratio2.before`
* Merchandise exports to GDP ratio average after NAFTA: `r ratio2.after`
* Merchandise imports to GDP ratio average before NAFTA: `r ratio3.before`
* Merchandise imports to GDP ratio average after NAFTA: `r ratio3.after`

\newpage

**iii.**

```{r services}
# Important services ratios
servExports <- goods_and_servicesExports - merchExports
servImports <- goods_and_servicesImports - merchImports
servExportsGDP <- t(servExports / gdp)
servImportsGDP <- t(servImports / gdp)
servGDP <- t((servExports  + servImports) / gdp)
services <- data.frame(servExportsGDP, servImportsGDP, servGDP)

# Mean before and after NAFTA
ratio4.before <- mean(servGDP[as.character(c(1960:1993)), ])
ratio4.after <- mean(servGDP[as.character(c(1994:2011)), ])

ratio5.before <- mean(servExportsGDP[as.character(c(1960:1993)), ])
ratio5.after <- mean(servExportsGDP[as.character(c(1994:2011)), ])

ratio6.before <- mean(servImportsGDP[as.character(c(1960:1993)), ])
ratio6.after <- mean(servImportsGDP[as.character(c(1994:2011)), ])

colours <-
  c(
    'Services trade to GDP' = '#051c2c',
    'Services exports to GDP' = '#1e40ca',
    'Services imports to GDP' = '#00a2ed'
  )

# Plot
ggplot(services) +
  geom_line(aes(
    x = as.numeric(row.names(services)),
    y = servGDP,
    colour = 'Services trade to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(services)),
    y = servExportsGDP,
    colour = 'Services exports to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(services)),
    y = servImportsGDP,
    colour = 'Services imports to GDP'
  )) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio4.before,
      yend = ratio4.before,
      colour = 'Services trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio4.after,
      yend = ratio4.after,
      colour = 'Services trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio5.before,
      yend = ratio5.before,
      colour = 'Services exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio5.after,
      yend = ratio5.after,
      colour = 'Services exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio6.before,
      yend = ratio6.before,
      colour = 'Services imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio6.after,
      yend = ratio6.after,
      colour = 'Services imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(x = 'year', y = 'ratio') +
  scale_colour_manual(name = 'Ratios', values = colours) +
  theme_minimal()
```

* Services trade to GDP ratio average before NAFTA: `r ratio4.before`
* Services trade to GDP ratio average after NAFTA: `r ratio4.after`
* Services exports to GDP ratio average before NAFTA: `r ratio5.before`
* Services exports to GDP ratio average after NAFTA: `r ratio5.after`
* Services imports to GDP ratio average before NAFTA: `r ratio6.before`
* Services imports to GDP ratio average after NAFTA: `r ratio6.after`

\newpage

## (b)

```{r data.2}
data.2 <- read.csv('data/WITS.csv') %>% subset(ProductDescription != '  UN Special Code')

exports.1993 <-
  data.2 %>% filter(Year == '1993' &
                      TradeFlowName == 'Export') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

imports.1993 <-
  data.2 %>% filter(Year == '1993' &
                      TradeFlowName == 'Import') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

exports.2007 <-
  data.2 %>% filter(Year == '2007' &
                      TradeFlowName == 'Export') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

imports.2007 <-
  data.2 %>% filter(Year == '2007' &
                      TradeFlowName == 'Import') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

exports.change <-
  merge(exports.1993, exports.2007, by = 'ProductDescription')
exports.change <-
  exports.change %>% add_column(
    perc.change = (
      exports.change$TradeValue.in.1000.USD.y / exports.change$TradeValue.in.1000.USD.x
    ) - 1
  ) %>% arrange(desc(perc.change))

imports.change <-
  merge(imports.1993, imports.2007, by = 'ProductDescription')
imports.change <-
  imports.change %>% add_column(
    perc.change = (
      imports.change$TradeValue.in.1000.USD.y / imports.change$TradeValue.in.1000.USD.x
    ) - 1
  )  %>% arrange(desc(perc.change))
```

**i.**

Top 10 exported commodities in 1993:
```{r}
kable(
  head(exports.1993, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl'
)
```

Top 10 exported commodities in 2007:
```{r}
kable(
  head(exports.2007, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl'
)
```

\newpage

**ii.**

Top 10 imported commodities in 1993:
```{r}
kable(
  head(imports.1993, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl'
)
```

Top 10 imported commodities in 2007:
```{r}
kable(
  head(imports.2007, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl'
)
```

\newpage

**iii.**

Top 10 commodities that experienced greatest growth in exports:
```{r}
kable(
  head(exports.change, 10),
  col.names = c('Product', 'Trade value in 1993', 'Trade value in 2007', 'Percentage change'),
  align = 'lrrr'
)
```

Top 10 commodities that experienced greatest growth in imports:
```{r}
kable(
  head(imports.change, 10),
  col.names = c('Product', 'Trade value in 1993', 'Trade value in 2007', 'Percentage change'),
  align = 'lrrr'
)
```

\newpage

**iv. and v.**

```{r 1993vpercChange, fig.height=10, fig.width=10}
colours <-
  c('Exports' = '#1e40ca',
    'Imports' = '#00a2ed')

ggplot() + geom_point(
  aes(
    x = exports.change$TradeValue.in.1000.USD.x,
    y = exports.change$perc.change,
    colour = 'Exports'
  )
) +
  geom_point(
    aes(
      x = imports.change$TradeValue.in.1000.USD.x,
      y = imports.change$perc.change,
      colour = 'Imports'
    )
  ) +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10') +
  labs(caption = '(log scale transform)',
       x = '1993',
       y = 'Percentage change') +
  scale_colour_manual(name = '', values = colours) +
  theme_minimal()
```

```{r}
corrExports.kendall <-
  cor(exports.change$TradeValue.in.1000.USD.x,
      exports.change$perc.change,
      method = 'kendall')
corrExports.pearson <-
  cor(exports.change$TradeValue.in.1000.USD.x,
      exports.change$perc.change,
      method = 'pearson')
corrExports.spearman <-
  cor(exports.change$TradeValue.in.1000.USD.x,
      exports.change$perc.change,
      method = 'spearman')

corrImports.kendall <-
  cor(imports.change$TradeValue.in.1000.USD.x,
      imports.change$perc.change,
      method = 'kendall')
corrImports.pearson <-
  cor(imports.change$TradeValue.in.1000.USD.x,
      imports.change$perc.change,
      method = 'pearson')
corrImports.spearman <-
  cor(imports.change$TradeValue.in.1000.USD.x,
      imports.change$perc.change,
      method = 'spearman')
```

There is almost *no linear relation* between data obtained from 1993 and their percentage change to 2007 as shown with Pearson's $r$ correlations. On the other hand, we can observe weakly **negative** monotonic and concordant relations with Spearman's $\rho$ and Kendall's $\tau$ coefficients.

|         | Pearson's $r$ | Spearman's $\rho$ | Kendall's $\tau$  |
|:--------|:-------------:|:-----------------:|:-----------------:|
| Exports | $`r corrExports.pearson`$ | $`r corrExports.spearman`$ | $`r corrExports.kendall`$ |
| Imports | $`r corrImports.pearson`$ | $`r corrImports.spearman`$ | $`r corrImports.kendall`$ |

\newpage

# Export and import wedges

```{r data.3}
data.3 <- read.csv('data/OECD.csv')

detrend <-
  function(country,
           index,
           date.1,
           date.2,
           date.3,
           date.4) {
    data <- data.3 %>% filter(LOCATION == index)
    
    GDP_CQRSA <- data %>%
      filter(SUBJECT == 'B1_GE',
             MEASURE == 'CQRSA') %>%
      subset(select = c('TIME', 'Value'))
    
    imports_CQRSA <- data %>%
      filter(SUBJECT == 'P7',
             MEASURE == 'CQRSA') %>%
      subset(select = c('TIME', 'Value'))
    
    GDP_DNBSA <- data %>%
      filter(SUBJECT == 'B1_GE',
             MEASURE == 'DNBSA') %>%
      subset(select = c('TIME', 'Value'))
    
    imports_DNBSA <- data %>%
      filter(SUBJECT == 'P7',
             MEASURE == 'DNBSA') %>%
      subset(select = c('TIME', 'Value'))
    
    # Data
    
    GDP <- data.frame(
      quarter = GDP_CQRSA$TIME,
      nominal = GDP_CQRSA$Value,
      real = GDP_CQRSA$Value / GDP_DNBSA$Value,
      deflactor = GDP_DNBSA$Value / 100
    )
    
    imports <- data.frame(
      quarter = imports_CQRSA$TIME,
      nominal = imports_CQRSA$Value,
      real = imports_CQRSA$Value / imports_DNBSA$Value,
      deflactor = imports_DNBSA$Value / 100
    )
    
    # log data
    
    log_GDP <- data.frame(
      quarter = GDP$quarter,
      nominal = log(GDP$nominal),
      real = log(GDP$real),
      deflactor = log(GDP$deflactor)
    )
    
    log_imports <- data.frame(
      quarter = imports$quarter,
      nominal = log(imports$nominal),
      real = log(imports$real),
      deflactor = log(imports$deflactor)
    )
    
    # Linear regression prediction
    logQ <- log_imports$real - log_GDP$real
    logX0 <- integer(length = length(logQ)) + 1
    logX1 <- GDP$deflactor - imports$deflactor
    
    X <- cbind(logX0, logX1)
    
    # Parameter estimation
    b <- inv(crossprod(X)) %*% (t(X) %*% logQ)
    
    # Prediction
    logQhat <- X %*% b
    log_imports$realPredicted <- logQhat + log_GDP$real
    
    # Import wedge
    log_imports$wedge <-
      log_imports$real - log_imports$realPredicted
    
    # Detrended data
    hp <-
      hpfilter(log_imports$wedge, freq = 1600, type = 'lambda')
    log_imports$trend <- hp$trend
    
    # Plot
    colours <- c('Import wedge' = '#1e40ca',
                 'Trend' = '#00a2ed')
    
    p <- ggplot() +
      geom_line(aes(
        x = as.Date(as.yearqtr(log_imports$quarter, format = '%Y-Q%q')),
        y = log_imports$wedge,
        group = 1,
        colour = 'Import wedge'
      )) +
      geom_line(aes(
        x = as.Date(as.yearqtr(log_imports$quarter, format = '%Y-Q%q')),
        y = log_imports$trend,
        group = 1,
        colour = 'Trend'
      )) +
      geom_rect(aes(
        xmin = as.Date(as.yearqtr(date.1, format = '%Y-Q%q')),
        xmax = as.Date(as.yearqtr(date.2, format = '%Y-Q%q')),
        ymin = min(log_imports$wedge),
        ymax = max(log_imports$wedge)
      ),
      alpha = 0.5) +
      geom_rect(aes(
        xmin = as.Date(as.yearqtr(date.3, format = '%Y-Q%q')),
        xmax = as.Date(as.yearqtr(date.4, format = '%Y-Q%q')),
        ymin = min(log_imports$wedge),
        ymax = max(log_imports$wedge)
      ),
      alpha = 0.5) +
      labs(title = country,
           x = 'Quarter',
           y = 'Percentage Change') +
      scale_x_date(
        guide = guide_axis(angle = 90),
        date_breaks = '1 year',
        date_labels = '%Y'
      ) +
      scale_colour_manual(name = '', values = colours) +
      theme_minimal()
    
    return(p)
  }
```

```{r importsCAN}
detrend(
    country = 'Canada',
    index = 'CAN',
    date.1 = '2001-Q2',
    date.2 = '2002-Q3',
    date.3 = '2008-Q4',
    date.4 = '2010-Q3'
)
```

```{r importsMEX}
detrend(
    country = 'Mexico',
    index = 'MEX',
    date.1 = '1994-Q4',
    date.2 = '1997-Q3',
    date.3 = '2008-Q3',
    date.4 = '2010-Q2'
)
```

```{r importsUSA}
detrend(
    country = 'United States of America',
    index = 'USA',
    date.1 = '2001-Q2',
    date.2 = '2004-Q2',
    date.3 = '2008-Q4',
    date.4 = '2010-Q4'
)
```

\newpage

# Heckscher–Ohlin

\newpage

```{r bib, include=FALSE}
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
