---
title: 'International Trade: Assignment 1'
runningheader: 'Assignment 1'
author: |
  Carlos Lezama, Marco Hinojosa, \
  Luis Jiménez and Antonio Carbonell
date: 'October 27, 2021'
output:
  tufte::tufte_handout:
    citation_package: natbib
header-includes:
  - \usepackage{pdfpages}
bibliography: references.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(tidyverse)
library(knitr)
library(expm)
library(matlib)
library(ggcorrplot)
library(mFilter)
library(zoo)

knitr::opts_chunk$set(
  cache.extra = packageVersion('tufte'),
  cache = FALSE,
  echo = FALSE,
  fig.fullwidth=TRUE,
  fig.height = 6,
  fig.pos = '!h',
  fig.width = 12,
  message = FALSE,
  out.extra = '',
  out.width = '100%',
  warning = FALSE
)

options(htmltools.dir.version = FALSE)
```

# Patterns of trade^[Data obtained from @databank and @wits.]

## (a)

**i. and ii.**

```{r data.1}
# Data import
data.1 <- read.csv('data/world-bank.csv') %>%
  slice(-c(6:10))
data.1 <- data.1[-c(1:4)]

# Columns rename
names(data.1) <- c(1960:2020)

# Variables
gdp <- data.1[1, ]
merchExports <- data.1[2, ]
merchImports <- data.1[3, ]
goods_and_servicesExports <- data.1[4, ]
goods_and_servicesImports <- data.1[5, ]
```

```{r merchandise}
# Important merchandise ratios
merchExportsGDP <- t(merchExports / gdp)
merchImportsGDP <- t(merchImports / gdp)
merchGDP <- t((merchExports  + merchImports) / gdp)
merchandise <- data.frame(merchExportsGDP, merchImportsGDP, merchGDP)

# Mean before and after NAFTA
ratio1.before <- mean(merchGDP[as.character(c(1960:1993)), ])
ratio1.after <- mean(merchGDP[as.character(c(1994:2011)), ])

ratio2.before <- mean(merchExportsGDP[as.character(c(1960:1993)), ])
ratio2.after <- mean(merchExportsGDP[as.character(c(1994:2011)), ])

ratio3.before <- mean(merchImportsGDP[as.character(c(1960:1993)), ])
ratio3.after <- mean(merchImportsGDP[as.character(c(1994:2011)), ])

colours <-
  c(
    'Merchandise trade to GDP' = '#051c2c',
    'Merchandise exports to GDP' = '#1e40ca',
    'Merchandise imports to GDP' = '#00a2ed'
  )

# Plot
ggplot(merchandise) +
  geom_line(aes(
    x = as.numeric(row.names(merchandise)),
    y = merchGDP,
    colour = 'Merchandise trade to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(merchandise)),
    y = merchExportsGDP,
    colour = 'Merchandise exports to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(merchandise)),
    y = merchImportsGDP,
    colour = 'Merchandise imports to GDP'
  )) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio1.before,
      yend = ratio1.before,
      colour = 'Merchandise trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio1.after,
      yend = ratio1.after,
      colour = 'Merchandise trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio2.before,
      yend = ratio2.before,
      colour = 'Merchandise exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio2.after,
      yend = ratio2.after,
      colour = 'Merchandise exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio3.before,
      yend = ratio3.before,
      colour = 'Merchandise imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio3.after,
      yend = ratio3.after,
      colour = 'Merchandise imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(x = 'year', y = 'ratio') +
  scale_colour_manual(name = 'Ratios', values = colours) +
  theme_minimal()
```

* Merchandise trade to GDP ratio average before NAFTA: `r ratio1.before`
* Merchandise trade to GDP ratio average after NAFTA: `r ratio1.after`
* Merchandise exports to GDP ratio average before NAFTA: `r ratio2.before`
* Merchandise exports to GDP ratio average after NAFTA: `r ratio2.after`
* Merchandise imports to GDP ratio average before NAFTA: `r ratio3.before`
* Merchandise imports to GDP ratio average after NAFTA: `r ratio3.after`

In all cases, the merchandise to GDP ratio increased such that trade activity increased for merchandise exports, imports and trade.

Furthermore, note that during the period from $1994$ to $2011$, the average increased more than $100\%$ for all merchandise trade. This means that the importance of trade increased significantly when comparing both periods.

\newpage

**iii.**

```{r services}
# Important services ratios
servExports <- goods_and_servicesExports - merchExports
servImports <- goods_and_servicesImports - merchImports
servExportsGDP <- t(servExports / gdp)
servImportsGDP <- t(servImports / gdp)
servGDP <- t((servExports  + servImports) / gdp)
services <- data.frame(servExportsGDP, servImportsGDP, servGDP)

# Mean before and after NAFTA
ratio4.before <- mean(servGDP[as.character(c(1960:1993)), ])
ratio4.after <- mean(servGDP[as.character(c(1994:2011)), ])

ratio5.before <- mean(servExportsGDP[as.character(c(1960:1993)), ])
ratio5.after <- mean(servExportsGDP[as.character(c(1994:2011)), ])

ratio6.before <- mean(servImportsGDP[as.character(c(1960:1993)), ])
ratio6.after <- mean(servImportsGDP[as.character(c(1994:2011)), ])

colours <-
  c(
    'Services trade to GDP' = '#051c2c',
    'Services exports to GDP' = '#1e40ca',
    'Services imports to GDP' = '#00a2ed'
  )

# Plot
ggplot(services) +
  geom_line(aes(
    x = as.numeric(row.names(services)),
    y = servGDP,
    colour = 'Services trade to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(services)),
    y = servExportsGDP,
    colour = 'Services exports to GDP'
  )) +
  geom_line(aes(
    x = as.numeric(row.names(services)),
    y = servImportsGDP,
    colour = 'Services imports to GDP'
  )) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio4.before,
      yend = ratio4.before,
      colour = 'Services trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio4.after,
      yend = ratio4.after,
      colour = 'Services trade to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio5.before,
      yend = ratio5.before,
      colour = 'Services exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio5.after,
      yend = ratio5.after,
      colour = 'Services exports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1960,
      xend = 1993,
      y = ratio6.before,
      yend = ratio6.before,
      colour = 'Services imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  geom_segment(
    aes(
      x = 1994,
      xend = 2011,
      y = ratio6.after,
      yend = ratio6.after,
      colour = 'Services imports to GDP'
    ),
    linetype = 'dotted'
  ) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  labs(x = 'year', y = 'ratio') +
  scale_colour_manual(name = 'Ratios', values = colours) +
  theme_minimal()
```

* Services trade to GDP ratio average before NAFTA: `r ratio4.before`
* Services trade to GDP ratio average after NAFTA: `r ratio4.after`
* Services exports to GDP ratio average before NAFTA: `r ratio5.before`
* Services exports to GDP ratio average after NAFTA: `r ratio5.after`
* Services imports to GDP ratio average before NAFTA: `r ratio6.before`
* Services imports to GDP ratio average after NAFTA: `r ratio6.after`

All exports, imports and trade in services relative to GDP ratios decreased after NAFTA, which means that services have a lesser impact on GDP. We cannot say that trade in services had decreased, but it is certain that it did not grow at the same pace as merchandise trade.

\newpage

## (b)

```{r data.2}
data.2 <- read.csv('data/WITS.csv') %>% subset(ProductDescription != '  UN Special Code')

exports.1993 <-
  data.2 %>% filter(Year == '1993' &
                      TradeFlowName == 'Export') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

imports.1993 <-
  data.2 %>% filter(Year == '1993' &
                      TradeFlowName == 'Import') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

exports.2007 <-
  data.2 %>% filter(Year == '2007' &
                      TradeFlowName == 'Export') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

imports.2007 <-
  data.2 %>% filter(Year == '2007' &
                      TradeFlowName == 'Import') %>% select(c('TradeValue.in.1000.USD', 'ProductDescription')) %>% arrange(desc(TradeValue.in.1000.USD))

exports.change <-
  merge(exports.1993, exports.2007, by = 'ProductDescription')
exports.change <-
  exports.change %>% add_column(
    perc.change = (
      exports.change$TradeValue.in.1000.USD.y / exports.change$TradeValue.in.1000.USD.x
    ) - 1
  ) %>% arrange(desc(perc.change))

imports.change <-
  merge(imports.1993, imports.2007, by = 'ProductDescription')
imports.change <-
  imports.change %>% add_column(
    perc.change = (
      imports.change$TradeValue.in.1000.USD.y / imports.change$TradeValue.in.1000.USD.x
    ) - 1
  )  %>% arrange(desc(perc.change))
```

**i.**

```{r}
kable(
  head(exports.1993, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl',
  caption = 'Top $10$ exported commodities in $1993$'
)
```

```{r}
kable(
  head(exports.2007, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl',
  caption = 'Top $10$ exported commodities in $2007$'
)
```

Each of the products included in the list increased its commercial value. Oil remained number one, even after NAFTA. The $2007$ list shows a preference for auto parts and electrical products, while the $1993$ list includes vegetables in $10$th place.

\newpage

**ii.**

```{r}
kable(
  head(imports.1993, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl',
  caption = 'Top $10$ imported commodities in $1993$'
)
```

```{r}
kable(
  head(imports.2007, 10),
  col.names = c('Trade value in $$1000$USD', 'Product'),
  align = 'cl',
  caption = 'Top $10$ imported commodities in $2007$'
)
```

Like exports, the trade value of the top $10$ imported commodities increased after NAFTA. The first place changed from valves to telecommunications equipment, but the composition remained substantially the same.

Clearly, preferences did not change, but the composition of exports did after economic openness or a greater focus on US and Canadian consumer preferences.

\newpage

**iii.**

```{r}
kable(
  head(exports.change, 10),
  col.names = c('Product', 'Trade value in $1993$', 'Trade value in $2007$', 'Percentage change'),
  align = 'lrrr',
  caption = 'Top $10$ commodities that experienced greatest growth in exports'
)
```

```{r}
kable(
  head(imports.change, 10),
  col.names = c('Product', 'Trade value in $1993$', 'Trade value in $2007$', 'Percentage change'),
  align = 'lrrr',
  caption = 'Top $10$ commodities that experienced greatest growth in imports'
)
```

\newpage

**iv. and v.**

```{r 1993vpercChange, fig.height=10, fig.width=10}
colours <-
  c('Exports' = '#1e40ca',
    'Imports' = '#00a2ed')

ggplot() + geom_point(
  aes(
    x = exports.change$TradeValue.in.1000.USD.x,
    y = exports.change$perc.change,
    colour = 'Exports'
  )
) +
  geom_point(
    aes(
      x = imports.change$TradeValue.in.1000.USD.x,
      y = imports.change$perc.change,
      colour = 'Imports'
    )
  ) +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10') +
  labs(caption = '(log scale)',
       x = '1993',
       y = 'Percentage change') +
  scale_colour_manual(name = '', values = colours) +
  theme_minimal()
```

```{r}
corrExports.kendall <-
  cor(exports.change$TradeValue.in.1000.USD.x,
      exports.change$perc.change,
      method = 'kendall')
corrExports.pearson <-
  cor(exports.change$TradeValue.in.1000.USD.x,
      exports.change$perc.change,
      method = 'pearson')
corrExports.spearman <-
  cor(exports.change$TradeValue.in.1000.USD.x,
      exports.change$perc.change,
      method = 'spearman')

corrImports.kendall <-
  cor(imports.change$TradeValue.in.1000.USD.x,
      imports.change$perc.change,
      method = 'kendall')
corrImports.pearson <-
  cor(imports.change$TradeValue.in.1000.USD.x,
      imports.change$perc.change,
      method = 'pearson')
corrImports.spearman <-
  cor(imports.change$TradeValue.in.1000.USD.x,
      imports.change$perc.change,
      method = 'spearman')
```

|         | Pearson's $r$ | Spearman's $\rho$ | Kendall's $\tau$  |
|:--------|:-------------:|:-----------------:|:-----------------:|
| Exports | $`r corrExports.pearson`$ | $`r corrExports.spearman`$ | $`r corrExports.kendall`$ |
| Imports | $`r corrImports.pearson`$ | $`r corrImports.spearman`$ | $`r corrImports.kendall`$ |

\newpage

There is almost *no linear relation* between data obtained from $1993$ and their percentage change to $2007$ as shown with Pearson's $r$ correlations. On the other hand, we can observe weakly **negative** monotonic and concordant relations with Spearman's $\rho$ and Kendall's $\tau$ coefficients.

\newpage

# Export and import wedges^[Data obtained from the @oecd]

```{r data.3}
data.3 <- read.csv('data/OECD.csv')

importsWedge <-
  function(country,
           date.1,
           date.2,
           date.3,
           date.4) {
    data <- data.3 %>% filter(Country == country)
    
    GDP_CQRSA <- data %>%
      filter(SUBJECT == 'B1_GE',
             MEASURE == 'CQRSA') %>%
      subset(select = c('TIME', 'Value'))
    
    imports_CQRSA <- data %>%
      filter(SUBJECT == 'P7',
             MEASURE == 'CQRSA') %>%
      subset(select = c('TIME', 'Value'))
    
    GDP_DNBSA <- data %>%
      filter(SUBJECT == 'B1_GE',
             MEASURE == 'DNBSA') %>%
      subset(select = c('TIME', 'Value'))
    
    imports_DNBSA <- data %>%
      filter(SUBJECT == 'P7',
             MEASURE == 'DNBSA') %>%
      subset(select = c('TIME', 'Value'))
    
    # Data
    
    GDP <- data.frame(
      quarter = GDP_CQRSA$TIME,
      nominal = GDP_CQRSA$Value,
      real = GDP_CQRSA$Value / GDP_DNBSA$Value,
      deflactor = GDP_DNBSA$Value / 100
    )
    
    imports <- data.frame(
      quarter = imports_CQRSA$TIME,
      nominal = imports_CQRSA$Value,
      real = imports_CQRSA$Value / imports_DNBSA$Value,
      deflactor = imports_DNBSA$Value / 100
    )
    
    # log data
    
    log_GDP <- data.frame(
      quarter = GDP$quarter,
      nominal = log(GDP$nominal),
      real = log(GDP$real),
      deflactor = log(GDP$deflactor)
    )
    
    log_imports <- data.frame(
      quarter = imports$quarter,
      nominal = log(imports$nominal),
      real = log(imports$real),
      deflactor = log(imports$deflactor)
    )
    
    # Linear regression prediction
    logQ <- log_imports$real - log_GDP$real
    logX0 <- integer(length = length(logQ)) + 1
    logX1 <- GDP$deflactor - imports$deflactor
    
    X <- cbind(logX0, logX1)
    
    # Parameter estimation
    b <- inv(crossprod(X)) %*% (t(X) %*% logQ)
    
    # Prediction
    logQhat <- X %*% b
    log_imports$realPredicted <- logQhat + log_GDP$real
    
    # Import wedge
    log_imports$wedge <-
      log_imports$real - log_imports$realPredicted
    
    # Detrended data
    hp <-
      hpfilter(log_imports$wedge, freq = 1600, type = 'lambda')
    log_imports$trend <- hp$trend
    
    # Plot
    colours <- c('Import wedge' = '#1e40ca',
                 'Trend' = '#00a2ed')
    
    p <- ggplot() +
      geom_line(aes(
        x = as.Date(as.yearqtr(log_imports$quarter, format = '%Y-Q%q')),
        y = log_imports$wedge,
        group = 1,
        colour = 'Import wedge'
      )) +
      geom_line(aes(
        x = as.Date(as.yearqtr(log_imports$quarter, format = '%Y-Q%q')),
        y = log_imports$trend,
        group = 1,
        colour = 'Trend'
      )) +
      geom_rect(aes(
        xmin = as.Date(as.yearqtr(date.1, format = '%Y-Q%q')),
        xmax = as.Date(as.yearqtr(date.2, format = '%Y-Q%q')),
        ymin = min(log_imports$wedge),
        ymax = max(log_imports$wedge)
      ),
      alpha = 0.5) +
      geom_rect(aes(
        xmin = as.Date(as.yearqtr(date.3, format = '%Y-Q%q')),
        xmax = as.Date(as.yearqtr(date.4, format = '%Y-Q%q')),
        ymin = min(log_imports$wedge),
        ymax = max(log_imports$wedge)
      ),
      alpha = 0.5) +
      labs(title = country,
           x = '',
           y = '') +
      scale_x_date(
        guide = guide_axis(angle = 90),
        date_breaks = '1 year',
        date_labels = '%Y'
      ) +
      scale_colour_manual(name = '', values = colours) +
      theme_minimal()
    
    return(p)
  }
```

```{r importsARG}
importsWedge(
    country = 'Argentina',
    date.1 = NA,
    date.2 = NA,
    date.3 = '2008-Q3',
    date.4 = '2010-Q2'
)
```

```{r importsCAN}
importsWedge(
    country = 'Canada',
    date.1 = '2001-Q2',
    date.2 = '2002-Q3',
    date.3 = '2008-Q4',
    date.4 = '2010-Q3'
)
```

```{r importsMEX}
importsWedge(
    country = 'Mexico',
    date.1 = '1994-Q4',
    date.2 = '1997-Q3',
    date.3 = '2008-Q3',
    date.4 = '2010-Q2'
)
```

```{r importsUSA}
importsWedge(
    country = 'United States',
    date.1 = '2001-Q2',
    date.2 = '2004-Q2',
    date.3 = '2008-Q4',
    date.4 = '2010-Q4'
)
```

\newpage

We computed the imports model prediction by the least squares method. Given

$$
x_{t,m}^{\text{obs}} - y_t = \alpha_m + \sigma (p_t - p_{t, m}),
$$

we can estimate

$$
\binom{\hat{\alpha}_m}{\hat{\sigma}} = \left( X^T X \right)^{-1} X^T Y
$$

where $X = \left[\mathbf{1}, p - p_m\right]_{t \times 2}$ and $Y = \left[ x_m^{\text{obs}} - y \right]_{t \times 1}$.

Thus, with our predictions $x_{t,m}^{\text{pred}} = X \cdot \left(\hat{\alpha}_m,\ \hat{\sigma}\right)^T + y_t$, import wedges were defined as follows:

$$
\tau_t = x_{t,m}^{\text{obs}} - x_{t,m}^{\text{pred}}.
$$

Plus, we used the following variables:

| $\log(\cdot)$ |   Data    |
|:-------------:|----------:|
| $p$   | GDP deflator      |
| $y$   | Volume of GDP     |
| $p_m$ | Imports deflator  |
| $x_m$ | Volume of imports |

> Remember that both the data and the estimate of $\alpha_m$, $\hat{\alpha}_m$, are scaled logarithmically. Furthermore, the volumes of GDP and imports were measured at constant prices.

Mexican and Argentinian economies show a similar behavior around the financial crisis of $2008$, since both present import wedges below the expected trend. On the other hand, it is interesting to analyze how Mexico's wedge in $1995$ is below the trend while Argentina's wedge in $2018$ is above its own. These two economies reacted differently towards their devaluations. It is also important to mention that the fall of the $1995$ import wedge of Mexico is a fall from an increasing trend, while Argentina increases in its wedge, it is a higher import wedge from a decreasing trend.

> See the **theoretical part** on the next page.

\includepdf[pages={-}]{Heckscher-Ohlin.pdf}

\newpage

# References
